Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2020-04-13 Paper/manuscript_v2/manuscript_07_01.tex
update Fig. 15 and title
Changed Normal to N
Notation

\title{Hilbert space approximate Bayesian \\ Gaussian processes using Stan}

@@

Thus, the Gram matrix $\mathrm{K}$ of the covariance function $k$ for a set of observations $i=1,\ldots,n$ and corresponding input values $\{x_i\}_{i=1}^{n} \in \Omega^{n}$ can be represented as
%
\begin{equation}
\mathrm{K} = \Phi \Delta \Phi^\intercal, \nonumber
\end{equation}
where $\Phi \in {\rm I\!R}^{n\times m}$ is the matrix of eigenfunctions $\phi_j(x_i)$
%
\begin{align}
\Phi =  \left[ {\begin{array}{ccc}
   \phi_1(x_1) & \cdots & \phi_m(x_1)  \\
    \vdots &\ddots & \vdots  \nonumber \\ 
    \phi_1(x_n) & \cdots & \phi_m(x_n) \\
  \end{array} } \right].
\end{align}
As a result, the model for $f$ can be written as
%
\begin{equation}
\bm{f} \sim \mathcal{N}(\bm{\mu},\Phi \Delta \Phi^\intercal). \nonumber
\end{equation}
This equivalently leads to a linear representation of $f$ via
%
\begin{equation}\label{eq_approxf}
f(x) \approx \sum_{j}^m \left( S_{\theta}(\sqrt{\lambda_j})\right)^{\frac{1}{2}} \phi_j(x) \beta_j,
\end{equation}
%
where $\beta_j \sim \text{Normal}(0,1)$. Thus, the function $f$ is approximated with a finite basis function expansion (using the eigenfunctions $\phi_j$ of the Laplace operator), scaled by the square root of spectral density values. A key property of this approximation is that the eigenfunctions $\phi_j$ do not depend on the hyperparameters of the covariance function $\theta$. Instead, the only dependence of the model on $\theta$ is through the spectral density $S_{\theta}$. The eigenvalues $\lambda_j$ are monotonically increasing with $j$ and $S_{\theta}$ goes rapidly to zero for bounded covariance functions. Therefore, eq.~\eqref{eq_approxf} can be expected to be a good approximation for a finite number of $m$ terms in the series as long as the inputs values $x_i$ are not too close to the boundaries $-L$ and $L$ of $\Omega$. The computational cost of evaluating the log posterior density of univariate HSGPs scales as $O(nm + m)$, where $n$ is the number of observations and $m$ the number of basis functions.

@@

Each multivariate eigenfunction $\phi^{\ast}_j$ corresponds to the product of the univariate eigenfunctions whose indices corresponds to the elements of the $D$-tuple $\mathbb{S}_{j\cdotp}$, and each multivariate eigenvalue $\bm{\lambda}^{\ast}_j$ is a $D$-vector with elements that are the univariate eigenvalues whose indices correspond to the elements of the $D$-tuple $\mathbb{S}_{j\bm{\cdotp}}$. Thus, for $\bm{x}=\{x_d\}_{d=1}^D \in \Omega$ and $j=1,2,\ldots,m^{\ast}$, we have 
%
\begin{align}
\bm{\lambda}^{\ast}_j &= \left\{ \lambda_{\mathbb{S}_{jd}} \right\}_{d=1}^D =  \left\{ \left(\frac{\pi \mathbb{S}_{jd}}{2L_d}\right)^{\!2} \right\}_{d=1}^D, \label{eq_eigenvalue_multi} \\
%
\phi^{\ast}_j(\bm{x}) &= \prod_{d=1}^{D} \phi_{\mathbb{S}_{jd}}(x_d) = \prod_{d=1}^{D} \sqrt{\frac{1}{L_d}} \sin\!\left(\sqrt{\lambda_{\mathbb{S}_{jd}}}(x_d+L_d)\right). \label{eq_eigenfunction_multi}
\end{align}
%
The approximate covariance function is then represented as
%
\begin{equation}\label{eq_approxcov_multi}
k(\bm{x},\bm{x}') \approx \sum_{j=1}^{m^{\ast}} 
S^{\ast}_{\theta}\left(\sqrt{\bm{\lambda}^{\ast}_j}\right)
\phi^{\ast}_j(\bm{x}) \phi^{\ast}_j(\bm{x}'),
\end{equation}
%
where $S^{\ast}_{\theta}$ is the spectral density of the $D$-dimensional covariance function (see Section \ref{ch4_sec_cov}). We can now write the approximate series expansion of the multivariate function $f$ as
%
\begin{equation}\label{eq_approxf_multi}
f(\bm{x}) \approx \sum_{j=1}^{m^{\ast}} 
\left( S^{\ast}_{\theta} \left(\sqrt{\bm{\lambda}^{\ast}_j} \right)\right)^{\! \frac{1}{2}} \phi^{\ast}_j(\bm{x}) \beta_j, 
\end{equation}
%
where, again, $\beta_j \sim \text{Normal}(0,1)$. The computational cost of evaluating the log posterior density of multivariate HSGPs scales as $O(n m^{\ast} + m^{\ast})$, where $n$ is the number of observations and $m^{\ast}$ is the number of multivariate basis functions. Although this still implies linear scaling in $n$, the approximation is more costly than in the univariate case, as $m^{\ast}$ is the product of the number of univariate basis functions over the input dimensions and grows exponentially with respect to the number of  dimensions.

@@

Comparing the accuracy of the lengthscale in Figure 
\ref{fig8_Tlscale_vs_Elscale} to the RMSE in Figure \ref{fig9_MSE_varing_lscale},
we see that they agree closely with each other for medium lengthscales.
That is, a good estimation of the lengthscale implies a
small RMSE. This is no longer true for very small or large lengthscales.
In small lengthscales, even very small inaccuracies 
may have a strong influence on the posteriors predictions and thus on the RMSE.
In large lengthscales, larger inaccuracies change the posterior predictions
only little and may thus not yield large RMSEs.
The dashed black line in Figure \ref{fig8_Tlscale_vs_Elscale} represents the 
minimum lengthscale that can be closely approximated under the given condition, according to the results presented in Figure \ref{fig6_lscale_vs_J_vs_c}. 
We observe that whenever the estimated lengthscale exceeds the minimally
estimable lengthscale, the RMSE of the posterior predictions is small
(see Figure \ref{fig9_MSE_varing_lscale}).
Conversely, when the estimated lengthscale is smaller than the minimally
estimable one, the RMSE becomes very large.

@@

This example consists of a simulated dataset with $n=250$ ($i=1,\dots,n$) single draws from a Gaussian process prior with a Mat{\'e}rn($\nu$=3/2) covariance function and hyperparameters marginal variance $\alpha=1$ and lengthscale $\ell=0.15$, with corresponding inputs values $\bm{x}=(x_1\dots,x_n)$ with $x_i \in [-1,1] \subset {\rm I\!R}$. To form the final noisy dataset $\bm{y}$, Gaussian noise standard deviation $\sigma=0.2$ was added to the GP draws.

The exact GP model for fitting this simulated dataset $\bm{y}$ can be written as follows,
%
\begin{align*}\label{ch5_eq_latentgp_simudata1}
\begin{split}
\bm{y} &= \bm{f} + \bm{\epsilon} \\
\bm{\epsilon} &\sim \mathcal{N}(0, \sigma^2  I) \\
f(x) &\sim \GP(0, k(x, x', \theta)),
\end{split}
\end{align*}
where $I$ represents the identity matrix and $\bm{f}=\{f(x_i)\}_{i=1}^n$ represents the underlying function values to the noisy data. The previous formulation corresponds to the latent form of a GP model. The function $f:{\rm I\!R} \to {\rm I\!R}$ is a GP prior with a Mat{\'e}rn($\nu$=3/2) covariance function $k$. Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $K$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,\dots,n$.

A more computationally efficient formulation of a GP model with Gaussian likelihood, and for probabilistic inference using MCMC sampling methods, would be its marginalized form,
%
\begin{equation*}\label{ch5_eq_marginalizedgp_simudata1}
\bm{y} \sim \mathcal{N}(0, K + \sigma^2 I ),
\end{equation*}
where the function values $\bm{f}$ have been integrated out, yielding a lower-dimensional parameter space over which to do inference, reducing the time of computation and improving the sampling and the effective number of samples.

In the HSGP model, the latent function values $f(x)$ are approximated as in eq.~(\ref{eq_approxf}), with the Mat{\'e}rn($\nu$=3/2) spectral density $S$ as in eq. ~(\ref{eq_specdens_32}), and eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ as in equations (\ref{eq_eigenvalue}) and (\ref{eq_eigenfunction}), respectively. 
In order to do model comparison, in addition to the exact GP model and HSGP model, a spline-based model is also fitted using the thin plate regression spline approach by \citet{wood2003thin} and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline model using the R-package \textit{brms} \citep{burkner2017brms}.

@@

The component $f_2(t)$ represents the yearly smooth seasonal pattern, using a periodic squared exponential covariance function (with period 365.25 to match the average length of the year) in a GP model,
%
\begin{equation*}
f_2(t) \sim \GP(0,k_2), \hspace{5mm} k_2(t,t') = \sigma_2^2 \exp\!\!\left(\!-\frac{2\,\text{sin}^{\!2}\!(\pi(t-t')/365.25}{\ell_2^2}\right).
\end{equation*}
% which means the function values $\bm{f}_2=\{f_2(t)\}_{t=1}^T$ are multivariate Gaussian distributed with covariance matrix $K_2$, where $K_{2_{t,s}}=k_2(t,s)$, with $t,s=1,\dots,T$.

@@

The function values of a GP model with this low-rank representation of the periodic exponential covariance function can be easily derived. Considering the identity
%
\begin{equation*}
\cos(j\omega_0 (x-x'))=\cos(j\omega_0 x) \cos(j\omega_0 x') + \sin(j\omega_0 x) \sin(j\omega_0 x'),
\end{equation*}
the covariance $k(\tau)$ in equation (\ref{ch5_eq_cov_periodic_taylor_approx}) can be written as
%
\begin{equation} \label{ch5_eq_cov_periodic_taylor_approx_2}
k(x,x')= \sigma^2 \left( \sum_{j=0}^{J} \tilde{q}_j^2 \cos(j\omega_0 x)  \cos(j\omega_0 x') + \sum_{j=1}^{J} \tilde{q}_j^2 \sin(j\omega_0 x) \sin(j\omega_0 x') \right).
\end{equation}
With this approximation for the periodic squared exponential covariance function $k(x,x')$, the approximate GP model $f(x) \sim \GP\left(0,k(x,x')\right)$ equivalently leads to a linear representation of $f(\cdot)$ via
%
\begin{equation} \label{ch5_eq_f_period}
f(x) \approx \sigma \left( \sum_{j=0}^J  \tilde{q}_j \cos(j\omega_0 x) \beta_j +  \sum_{j=1}^J \tilde{q}_j \sin(j\omega_0 x) \beta_{J+1+j} \right),
\end{equation}
where $\beta_j \sim \text{Normal}(0,1)$, with $j=1,\dots,2J+1$. The cosine $\cos(j\omega_0 x)$ and sinus $\sin(j\omega_0 x)$ terms do not depend on the covariance hyperparameters $\ell$. The only dependence on the hyperparameter $\ell$ is through the coefficients $\tilde{q}_j$, which are $J$-dimensional. The computational cost of this approximation scales as $O\big(n(2J+1) + (2J+1)\big)$, where $n$ is the number of observations and $J$ the number of cosine terms.
The parameterization in equation (\ref{ch5_eq_f_period}) is naturally in the non-centered form with independent prior distributions on
$\beta_j$, which makes posterior inference easier.

>>>>

\title{Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming}

@@

Thus, the Gram matrix $\mathrm{K}$ of the covariance function $k$ for a set of observations $i=1,\ldots,n$ and corresponding input values $\{x_i\}_{i=1}^{n} \in \Omega^{n}$ can be represented as
%
\begin{equation}
\mathrm{K} = \Phi \Delta \Phi^\intercal, \nonumber
\end{equation}
where $\Phi \in {\rm I\!R}^{n\times m}$ is the matrix of eigenfunctions $\phi_j(x_i)$
%
\begin{align}
\Phi =  \left[ {\begin{array}{ccc}
   \phi_1(x_1) & \cdots & \phi_m(x_1)  \\
    \vdots &\ddots & \vdots  \nonumber \\ 
    \phi_1(x_n) & \cdots & \phi_m(x_n) \\
  \end{array} } \right].
\end{align}
As a result, the model for $f$ can be written as
%
\begin{equation}
\bm{f} \sim \mathcal{N}(\bm{\mu},\Phi \Delta \Phi^\intercal). \nonumber
\end{equation}
This equivalently leads to a linear representation of $f$ via
%
\begin{equation}\label{eq_approxf}
f(x) \approx \sum_{j}^m \left( S_{\theta}(\sqrt{\lambda_j})\right)^{\frac{1}{2}} \phi_j(x) \beta_j,
\end{equation}
%
where $\beta_j \sim \mathcal{N}(0,1)$. Thus, the function $f$ is approximated with a finite basis function expansion (using the eigenfunctions $\phi_j$ of the Laplace operator), scaled by the square root of spectral density values. A key property of this approximation is that the eigenfunctions $\phi_j$ do not depend on the hyperparameters of the covariance function $\theta$. Instead, the only dependence of the model on $\theta$ is through the spectral density $S_{\theta}$. The eigenvalues $\lambda_j$ are monotonically increasing with $j$ and $S_{\theta}$ goes rapidly to zero for bounded covariance functions. Therefore, eq.~\eqref{eq_approxf} can be expected to be a good approximation for a finite number of $m$ terms in the series as long as the inputs values $x_i$ are not too close to the boundaries $-L$ and $L$ of $\Omega$. The computational cost of evaluating the log posterior density of univariate HSGPs scales as $O(nm + m)$, where $n$ is the number of observations and $m$ the number of basis functions.

@@

Each multivariate eigenfunction $\phi^{\ast}_j$ corresponds to the product of the univariate eigenfunctions whose indices corresponds to the elements of the $D$-tuple $\mathbb{S}_{j\cdotp}$, and each multivariate eigenvalue $\bm{\lambda}^{\ast}_j$ is a $D$-vector with elements that are the univariate eigenvalues whose indices correspond to the elements of the $D$-tuple $\mathbb{S}_{j\bm{\cdotp}}$. Thus, for $\bm{x}=\{x_d\}_{d=1}^D \in \Omega$ and $j=1,2,\ldots,m^{\ast}$, we have 
%
\begin{align}
\bm{\lambda}^{\ast}_j &= \left\{ \lambda_{\mathbb{S}_{jd}} \right\}_{d=1}^D =  \left\{ \left(\frac{\pi \mathbb{S}_{jd}}{2L_d}\right)^{\!2} \right\}_{d=1}^D, \label{eq_eigenvalue_multi} \\
%
\phi^{\ast}_j(\bm{x}) &= \prod_{d=1}^{D} \phi_{\mathbb{S}_{jd}}(x_d) = \prod_{d=1}^{D} \sqrt{\frac{1}{L_d}} \sin\!\left(\sqrt{\lambda_{\mathbb{S}_{jd}}}(x_d+L_d)\right). \label{eq_eigenfunction_multi}
\end{align}
%
The approximate covariance function is then represented as
%
\begin{equation}\label{eq_approxcov_multi}
k(\bm{x},\bm{x}') \approx \sum_{j=1}^{m^{\ast}} 
S^{\ast}_{\theta}\left(\sqrt{\bm{\lambda}^{\ast}_j}\right)
\phi^{\ast}_j(\bm{x}) \phi^{\ast}_j(\bm{x}'),
\end{equation}
%
where $S^{\ast}_{\theta}$ is the spectral density of the $D$-dimensional covariance function (see Section \ref{ch4_sec_cov}). We can now write the approximate series expansion of the multivariate function $f$ as
%
\begin{equation}\label{eq_approxf_multi}
f(\bm{x}) \approx \sum_{j=1}^{m^{\ast}} 
\left( S^{\ast}_{\theta} \left(\sqrt{\bm{\lambda}^{\ast}_j} \right)\right)^{\! \frac{1}{2}} \phi^{\ast}_j(\bm{x}) \beta_j, 
\end{equation}
%
where, again, $\beta_j \sim \mathcal{N}(0,1)$. The computational cost of evaluating the log posterior density of multivariate HSGPs scales as $O(n m^{\ast} + m^{\ast})$, where $n$ is the number of observations and $m^{\ast}$ is the number of multivariate basis functions. Although this still implies linear scaling in $n$, the approximation is more costly than in the univariate case, as $m^{\ast}$ is the product of the number of univariate basis functions over the input dimensions and grows exponentially with respect to the number of  dimensions.

@@

Comparing the accuracy of the lengthscale in Figure~\ref{fig8_Tlscale_vs_Elscale} to the RMSE in Figure~\ref{fig9_MSE_varing_lscale}, we see that they agree closely with each other for medium lengthscales. That is, a good estimation of the lengthscale implies a small RMSE. This is no longer true for very small or large lengthscales. In small lengthscales, even very small inaccuracies  may have a strong influence on the posteriors predictions and thus on the RMSE. In large lengthscales, larger inaccuracies change the posterior predictions only little and may thus not yield large RMSEs. The dashed black line in Figure \ref{fig8_Tlscale_vs_Elscale} represents the  minimum lengthscale that can be closely approximated under the given condition, according to the results presented in Figure \ref{fig6_lscale_vs_J_vs_c}.  We observe that whenever the estimated lengthscale exceeds the minimally estimable lengthscale, the RMSE of the posterior predictions is small
(see Figure \ref{fig9_MSE_varing_lscale}). Conversely, when the estimated lengthscale is smaller than the minimally estimable one, the RMSE becomes very large.

@@

This example consists of a simulated dataset with $n=250$ ($i=1,2,\dots,n$) single draws from a Gaussian process prior with a Mat{\'e}rn($\nu$=3/2) covariance function and hyperparameters marginal variance $\alpha=1$ and lengthscale $\ell=0.15$, with corresponding inputs values $\bm{x}=(x_1,x_2,\dots,x_n)$ with $x_i \in [-1,1] \subset {\rm I\!R}$. To form the final noisy dataset $\bm{y}$, Gaussian noise standard deviation $\sigma=0.2$ was added to the GP draws.

The exact GP model for fitting this simulated dataset $\bm{y}$ can be written as follows,
%
\begin{align*}\label{ch5_eq_latentgp_simudata1}
\begin{split}
\bm{y} &= \bm{f} + \bm{\epsilon} \\
\bm{\epsilon} &\sim \mathcal{N}(0, \sigma^2  I) \\
f(x) &\sim \GP(0, k(x, x', \theta)),
\end{split}
\end{align*}
%
where $I$ represents the identity matrix and $\bm{f}=\{f(x_i)\}_{i=1}^n$ represents the underlying function values to the noisy data. The previous formulation corresponds to the latent form of a GP model. The function $f:{\rm I\!R} \to {\rm I\!R}$ is a GP prior with a Mat{\'e}rn($\nu$=3/2) covariance function $k$. Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $K$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,\dots,n$.

A more computationally efficient formulation of a GP model with Gaussian likelihood, and for probabilistic inference using MCMC sampling methods, would be its marginalized form,
%
\begin{equation*}\label{ch5_eq_marginalizedgp_simudata1}
\bm{y} \sim \mathcal{N}(0, K + \sigma^2 I ),
\end{equation*}
%
where the function values $\bm{f}$ have been integrated out, yielding a lower-dimensional parameter space over which to do inference, reducing the time of computation and improving the sampling and the effective number of samples.

In the HSGP model, the latent function values $f(x)$ are approximated as in eq.~(\ref{eq_approxf}), with the Mat{\'e}rn($\nu$=3/2) spectral density $S$ as in eq.~(\ref{eq_specdens_32}), and eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ as in equations (\ref{eq_eigenvalue}) and (\ref{eq_eigenfunction}), respectively.  In order to do model comparison, in addition to the exact GP model and HSGP model, a spline-based model is also fitted using the thin plate regression spline approach by \citet{wood2003thin} and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline model using the R-package \textit{brms} \citep{burkner2017brms}.

@@

The component $f_2(t)$ represents the yearly smooth seasonal pattern, using a periodic squared exponential covariance function (with period 365.25 to match the average length of the year) in a GP model,
%
\begin{equation*}
f_2(t) \sim \GP(0,k_2), \hspace{5mm} k_2(t,t') = \sigma_2^2 \exp\!\!\left(\!-\frac{2\,\mathrm{sin}^{\!2}\!(\pi(t-t')/365.25}{\ell_2^2}\right).
\end{equation*}
% which means the function values $\bm{f}_2=\{f_2(t)\}_{t=1}^T$ are multivariate Gaussian distributed with covariance matrix $K_2$, where $K_{2_{t,s}}=k_2(t,s)$, with $t,s=1,\dots,T$.

@@

The function values of a GP model with this low-rank representation of the periodic exponential covariance function can be easily derived. Considering the identity
%
\begin{equation*}
\cos(j\omega_0 (x-x'))=\cos(j\omega_0 x) \cos(j\omega_0 x') + \sin(j\omega_0 x) \sin(j\omega_0 x'),
\end{equation*}
the covariance $k(\tau)$ in equation (\ref{ch5_eq_cov_periodic_taylor_approx}) can be written as
%
\begin{equation} \label{ch5_eq_cov_periodic_taylor_approx_2}
k(x,x')= \sigma^2 \left( \sum_{j=0}^{J} \tilde{q}_j^2 \cos(j\omega_0 x)  \cos(j\omega_0 x') + \sum_{j=1}^{J} \tilde{q}_j^2 \sin(j\omega_0 x) \sin(j\omega_0 x') \right).
\end{equation}
With this approximation for the periodic squared exponential covariance function $k(x,x')$, the approximate GP model $f(x) \sim \GP\left(0,k(x,x')\right)$ equivalently leads to a linear representation of $f(\cdot)$ via
%
\begin{equation} \label{ch5_eq_f_period}
f(x) \approx \sigma \left( \sum_{j=0}^J  \tilde{q}_j \cos(j\omega_0 x) \beta_j +  \sum_{j=1}^J \tilde{q}_j \sin(j\omega_0 x) \beta_{J+1+j} \right),
\end{equation}
where $\beta_j \sim \mathcal{N}(0,1)$, with $j=1,\dots,2J+1$. The cosine $\cos(j\omega_0 x)$ and sinus $\sin(j\omega_0 x)$ terms do not depend on the covariance hyperparameters $\ell$. The only dependence on the hyperparameter $\ell$ is through the coefficients $\tilde{q}_j$, which are $J$-dimensional. The computational cost of this approximation scales as $O\big(n(2J+1) + (2J+1)\big)$, where $n$ is the number of observations and $J$ the number of cosine terms.
The parameterization in equation (\ref{ch5_eq_f_period}) is naturally in the non-centered form with independent prior distributions on
$\beta_j$, which makes posterior inference easier.
